---
title: "Group AA Milestone 6"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: '2022-12-11'
group members: Victoria, Danni, Maddy 
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```
### Problem statement

Smoking cigarettes can damage the body's organs, increase the risk of numerous illnesses, and generally harm people's health. The most common preventable cause of mortality in the U.S. is smoking cigarettes. Smoking is linked to numerous cancers, heart conditions, strokes, respiratory conditions such chronic obstructive pulmonary disease, diabetes, and other disorders. However, some smokers are more vulnerable to these negative effects than others. The question we explored is "what characteristics and behaviors among smokers in California lead to adverse health outcomes?"

At any age, even among heavy and longtime smokers, quitting smoking can have a significant positive impact on both short- and long-term health. For the prevention and cessation of smoking, a range of therapies are successful. With the use of data from the 2011 California Smokers' Cohort (CSC), the California Tobacco Surveys (CTS), we may better understand the tobacco use and smoking habits of Californian smokers and put initiatives into place in high-risk areas to encourage quitting.

### Methods

In the 2011 CSC study, 1,000 smokers completed the follow up survey and were included in the data set for analysis. We cleaned the data for analysis of 5 selected health outcomes, as well as select demographic variables. Our data were originally split in two data sets, which we combined by merging on the patient ID variable. Once that was finalized we cleaned and analyzed the data by factoring all variables that were characters, combining all race variables into one column, accounting for missing variables, and calculating pack years. The study population did not have a very diverse racial/ethnic spread, so we decided to exclude race in our analysis. We decided to focus on health outcomes and the stratification by income level, and subset our data accordingly to include variables that would be of importance in our visualizations and analysis. We also created new variables to categorize income into low, medium, and high income, as well as one to categorize pack years to "average", "above average", and "below average" pack years. 

```{r importing demographic data, include=FALSE}
#import CA Smoker data set

install.packages("ggpubr")
library("ggpubr")
library("stringr")

ca_smoker_info <- read_csv("~/PHW251_Fall2022/phw251_projectdata/ca_csc_smoker_data.csv")

ca_smoker_info$HOWMANY <- as.character(ca_smoker_info$HOWMANY)

#tidying data
ca_smoker_selected <- ca_smoker_info %>% select(c(psraid,smokstat,HOWMANY, SMOK6NUM, SMOK6UNI)) %>% 
rename(ID = psraid, smoking_status = smokstat, howmany = HOWMANY, smok6num = SMOK6NUM, smok6uni = SMOK6UNI) 

ca_smoker_selected$howmany <- as.numeric(ca_smoker_selected$howmany)
ca_smoker_selected$smok6num <- as.numeric(ca_smoker_selected$smok6num)

ca_smoker_selected <- ca_smoker_selected %>%
  mutate(pack_year = howmany/20*smok6num)
```

```{r importing outcome data, include=FALSE}
#import CA smoker disease outcome and race data set
ca_outcome_race <- read_csv("~/PHW251_Fall2022/phw251_projectdata/ca_csc_outcome_race_data.csv")

#tidying data
ca_outcome_race_selected <- ca_outcome_race %>% select(c(ID, SOCIAL, ASTHMA, 
                                                         HEARTDIS, DIABETES, 
                                                         OTHMENILL, INCOME, 
                                                         race01, race02, race03, 
                                                         race04, race05, race06, 
                                                         race07, race08, race09, 
                                                         race10, race11, race12, 
                                                         race13,race14, race15)) %>%
rename (social = SOCIAL, asthma = ASTHMA, heartdis = HEARTDIS, diabetes = DIABETES, 
        othmenill = OTHMENILL, income = INCOME)

```

```{r, include=FALSE}
#remove "DIS" & "STAT" in the "ID" column 
ca_outcome_race_selected$ID <- gsub("DIS","",as.character(ca_outcome_race_selected$ID))
ca_outcome_race_selected$ID <- gsub("STAT","",as.character(ca_outcome_race_selected$ID))

#joining two data sets by participant's unique ID
ca_smoker_outcome <- merge(x = ca_smoker_selected, y = ca_outcome_race_selected, by = "ID")

```

```{r create single race variable, include=FALSE}

#use mutate to combine 15 binary race columns into one categorical variable called "race"

ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(race = case_when(race01 == "Yes" ~ "White",
                          race02 == "Yes" ~ "Black",
                          race03 == "Yes" ~ "Japanese",
                          race04 == "Yes" ~ "Chinese",
                          race05 == "Yes" ~ "Filipino",
                          race06 == "Yes" ~ "Korean",
                          race12 == "Yes" ~ "Vietnamese",
                          race07 == "Yes" ~ "Other Asian Pacific Islander",
                          race08 == "Yes" ~ "American Indian Alaska Native",
                          race09 == "Yes" ~ "Mexican",
                          race10 == "Yes" ~ "Hispanic or Latino",
                          race11 == "Yes" ~ "Other",
                          race13 == "Yes" ~ "Asian Indian",
                          race14 == "Yes" ~ "Refused",
                          race15 == "Yes" ~ "Don't Know"))

#drop leftover binary race columns 

ca_smoker_outcome <- select(ca_smoker_outcome,-race01,
                                 -race02,
                                 -race03,
                                 -race04,
                                 -race05,
                                 -race06,
                                 -race07,
                                 -race08,
                                 -race09,
                                 -race10,
                                 -race11,
                                 -race12,
                                 -race13,
                                 -race14,
                                 -race15)
                            

```

```{r, include=FALSE}
#identify types of each data element
str(ca_smoker_outcome)

#convert data types to appropriate type in new column such as as.factor; as.numeric; as.character
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(new_howmany = as.numeric(howmany)) %>%
  mutate(new_smoking_status= as.factor(smoking_status)) %>% 
  mutate(new_smok6num = as.numeric(smok6num)) %>% 
  mutate(new_smok6uni = as.factor(smok6uni)) %>%
  mutate(new_pack_year = as.numeric(pack_year)) %>% 
  mutate(new_social = as.factor(social)) %>% 
  mutate(new_asthma = as.factor(asthma)) %>% 
  mutate(new_heartdis = as.factor(heartdis)) %>% 
  mutate(new_diabetes = as.factor(heartdis)) %>% 
  mutate(new_othmenill = as.factor(othmenill)) %>% 
  mutate(new_income = as.factor(income))

summary(ca_smoker_outcome$new_howmany)
```

```{r explore smoking status variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(smoking_status))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(smoking_status))

#tabulate smoking status
table(ca_smoker_outcome$smoking_status)
```

```{r explore race variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(race))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(race))

#tabulate smoking status
table(ca_smoker_outcome$race)
```

```{r explore income variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(income))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(income))

#tabulate income
table(ca_smoker_outcome$income)
```

```{r explore heart disease outcome variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(heartdis))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(heartdis))

#tabulate income
table(ca_smoker_outcome$heartdis)
```

```{r explore pack years variable, include=FALSE}

#look at minimum, median, mean, maximum, and # of NAs in pack year
summary(ca_smoker_outcome$new_pack_year)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)

is.na(ca_smoker_outcome) <- ca_smoker_outcome == "(DO NOT READ) Don't know"
is.na(ca_smoker_outcome) <- ca_smoker_outcome == "(DO NOT READ) Refused"
is.na(ca_smoker_outcome) <- ca_smoker_outcome == "NA/Not Applicable"

#subsetting rows  and columns to include only "new" columns, rows include NA 

str(ca_smoker_outcome)
ca_smoker_outcome <- select(ca_smoker_outcome, 
                            c('ID', 'new_smoking_status', 'new_howmany', 
                              'new_smok6num', 'new_smok6uni', 'new_pack_year',
                              'new_social', 'new_asthma', 'new_heartdis',
                              'new_diabetes', 'new_othmenill', 'new_income'))
```

```{r, include=FALSE}
library(stringr)
ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(smok_daily = str_c(ca_smoker_outcome$new_smok6num," ",
                            ca_smoker_outcome$new_smok6nuni)) 
```

```{r, include=FALSE}
library(dplyr)
summary(ca_smoker_outcome$new_pack_year)
ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(pack_year_avg = mean(ca_smoker_outcome$new_pack_year, na.rm = TRUE))
ca_smoker_outcome$pack_year_avg <- round(ca_smoker_outcome$pack_year_avg ,digit= 0)
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(pack_year_avg_level = case_when
                                                  (new_pack_year > pack_year_avg ~ "above average", 
                                                    new_pack_year < pack_year_avg ~ "below average",
                                                    new_pack_year == pack_year_avg ~ "average"))

```

```{r, include=FALSE}
unique(ca_smoker_outcome$new_income)
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(income_levels = case_when( 
  new_income %in% c("$20,000 or less", "$20,001 to $30,000") ~ "Low Income",
new_income %in% c("$30,001 to $50,000", "$50,001 to $75,000") ~ "Middle Income",
new_income %in% c("$75,001 to $100,000", "$100,001 to $150,000", "Over $150,000") ~ "High Income")) 
```

```{r, include=FALSE}
table(ca_smoker_outcome$new_smoking_status, useNA = "always")
table(ca_smoker_outcome$new_smok6uni, useNA = "always")
table(ca_smoker_outcome$new_social, useNA = "always")
table(ca_smoker_outcome$new_asthma, useNA = "always")
table(ca_smoker_outcome$new_heartdis, useNA = "always")
table(ca_smoker_outcome$new_diabetes, useNA = "always")
table(ca_smoker_outcome$new_othmenill, useNA = "always")
table(ca_smoker_outcome$new_income, useNA = "always")
```

```{r other new variable brainstorm, include=FALSE}

head(ca_smoker_outcome)

#if in years, leave as is
#if in months, divide by 12
#could add new column, conditional = 1 if units is in years, equals 12 if units is in months
#then create another new column for daily smoking length in years
#then update formula for pack years to reflect this

test_variable_df <- ca_smoker_outcome %>%
  mutate(temp_var = case_when(new_smok6uni=="Months" ~ 12,
                              new_smok6uni=="Years" ~ 1, 
                              new_smok6uni=="Days" ~ 365))

test_variable_df <- test_variable_df %>% 
  mutate(new_smok6num_inyears = new_smok6num/temp_var)

```

\newpage

### Visualizations

#### Visual 1: Table One of Pack Years and Health Outcomes

```{r, include=FALSE}
library(kableExtra)
library(tableone) #load table one library

#build new dataset with only necessary variables
visual3_df <- test_variable_df %>% 
  select(new_asthma, new_diabetes, new_heartdis, new_pack_year, new_othmenill, pack_year_avg_level) %>%
  rename(Asthma=new_asthma,
         "Above/Below Average Pack Years" = pack_year_avg_level,
         Diabetes=new_diabetes, 
         "Heart Disease"=new_heartdis, 
         "Pack Years"=new_pack_year, 
         "Other Mental Health" =new_othmenill)
#create table one
visual3table <- CreateTableOne(data=visual3_df,
  vars=c("Pack Years", "Asthma", "Diabetes", "Heart Disease", "Other Mental Health"),
  factorVars = c("Asthma", "Diabetes", "Heart Disease", "Other Mental Health", "Above/Below Average Pack Years"),
  strata="Above/Below Average Pack Years")

```

```{r, echo=FALSE}
kable(print(visual3table, showAllLevels=TRUE), 
      caption="Pack Years and Health Outcomes Stratified by Above/Below Average Pack Years")
```

```{r hiding 1 interpretations for final report in favor of a results section, include=FALSE}
#Interpretation: 
# This table is a "table one": a common table of descriptive statistics for a study sample. This table one displays descriptive statistics for our health outcomes of interest, as well as the pack years variable. All are stratified by one of our factor variables, "above/below pack years average", which displays whether the participant's pack years calculated in Milestone 3 were above, at, or below the average pack years of the entire sample. 
```
\newpage

#### Visual 2: Outcome and Income Cross Tabulations 

```{r, include=FALSE}
library(janitor)
library(kableExtra)
library(tidyverse)
```

```{r, echo=FALSE}
# For asthma status data set
a <- test_variable_df %>%
  tabyl(new_asthma, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Asthma_status = new_asthma)

a <- a[, c("Asthma_status", "High Income", "Middle Income", "Low Income", "Total")]

a %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 1, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Asthma Status x Income Level", 5))
```

```{r hiding 2 interpretations for final report in favor of a results section, include=FALSE}

# Interpretation: 
# Of the 176 patients with asthma, 30 were reported as the high-income group, 49 as the middle-income group, and 97 as the low-income group. 
# Of the 762 study participants without asthma, 187 were reported as the high-income group, 290 as the middle-income group, and 285 as the low-income group.  

```



```{r, echo=FALSE}
# For heart disease status data set
b <- test_variable_df %>%
  tabyl(new_heartdis, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Heartdis_status = new_heartdis)

b <- b[-1, c("Heartdis_status", "High Income", "Middle Income", "Low Income", "Total")]

b %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 2, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Heart Disease Status x Income Level", 6))
```

```{r 3 hiding interpretations for final report in favor of a results section, include=FALSE}

# Interpretation:
# Of the 77 patients with heart disease, 14 were reported as the high-income group, 20 as the middle-income group, and 43 as the low-income group. 
# Of the 858 study participants without heart disease, 202 were reported as the high-income group, 319 as the middle-income group, and 337 as the low-income group.

```

```{r, echo=FALSE}
# For diabetes status data set
c <- test_variable_df %>%
  tabyl(new_diabetes, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Diabetes_status = new_diabetes)

c <- c[-1, c("Diabetes_status", "High Income", "Middle Income", "Low Income", "Total")]

c %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 2, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Diabetes Status x Income Level", 6))
```

```{r 4 hiding interpretations for final report in favor of a results section, include=FALSE}

#Interpretation:

# Of the 77 patients with heart disease, 14 were reported as the high-income group, 20 as the middle-income group, and 43 as the low-income group. 
# Of the 858 study participants without heart disease, 202 were reported as the high-income group, 319 as the middle-income group, and 337 as the low-income group. 

```

\newpage

#### Visual 3: Of those who do have health outcomes, how are they distributed by income status. How are those that do not have health outcomes distributed by income?

```{r, include=FALSE}
library(ggplot2)
# Changed New variables to factors from most recent data set 
test_variable_df$pack_year_avg_level <- as.factor(test_variable_df$pack_year_avg_level)
test_variable_df$income_levels <- factor(test_variable_df$income_levels)
test_variable_df$smok_daily <- as.numeric(test_variable_df$smok_daily)
str(test_variable_df)

#Adding levels to Income Level 
test_variable_df$income_levels <- ordered(test_variable_df$income_levels, levels = c("Low Income", "Middle Income", "High Income"))
levels(test_variable_df$income_levels)

#Adding levels to pack year average level
test_variable_df$pack_year_avg_level <- ordered(test_variable_df$pack_year_avg_level, levels = c("below average", "average", "above average"))
levels(test_variable_df$pack_year_avg_level)
```

```{r Asthma data set and ggplot, echo=FALSE}
asthma_by_income <- test_variable_df %>% select(new_asthma, income_levels) %>% 
  group_by(new_asthma, income_levels) %>%  mutate(asthma_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

A <- ggplot(data = na.omit(asthma_by_income), aes(x = new_asthma, y= asthma_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
    scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Asthama", y = "# of Patients",
       title = "Patients and Their Asthma Status Distributed by Income Status", 
       caption = "Note: Patients who did not respond were excluded")
```

```{r Heart Disease data set and ggplot, echo=FALSE}
heartdis_by_income <- test_variable_df %>% select(new_heartdis, income_levels) %>% 
  group_by(new_heartdis, income_levels) %>%  mutate(heartdis_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

B <- ggplot(data = heartdis_by_income, aes(x = new_heartdis, y= heartdis_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
  scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Heart Disease", y = "# of Patients", 
       title = "Patients and Their Heart Disease Status Distributed by Income Status",
       caption = "Note: Patients who did not respond were excluded")

```

```{r Diabetes data set and ggplot, echo=FALSE}
diabetes_by_income <- test_variable_df %>% select(new_diabetes, income_levels) %>% 
  group_by(new_diabetes, income_levels) %>%  mutate(diabetes_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

C<- ggplot(data = diabetes_by_income, aes(x = new_diabetes, y= diabetes_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
  scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Diabetes", y = "# of Patients", 
       title = "Patients and Their Diabetes Status Distributed by Income Status",
       caption = "Note: Patients who did not respond were excluded")
 
```

Prevalence of Asthma, Heart Disease and Diabetes Stratified by Income Status

```{r, include=FALSE}
# The following code allows download of ggpubr to utilize the ggarrange function to 
# combine all separate ggplots together. 

if(!require(devtools)) install.packages("devtools")
devtools::install_github("kassambara/ggpubr")
install.packages("ggpubr")
library(ggpubr)
```

```{r, echo=FALSE}
# Hidden code allows download of ggpubr to utilize the ggarrange function to 
# combine all separate ggplots together. 

figure_one <- ggarrange(A, B, C, ncol = 1, nrow = 3)
figure_one
```

```{r 5 hiding interpretations for final report in favor of a results section, include=FALSE}

# Interpretation:
# 
# In Patients and Their Asthma Status Distributed by Income Status, of the 762 patients that reported no asthma, 285 patients came from low-income households, 290 patients came from middle-income households, and 187 patients came from high-income households. Of the 176 patients that reported having asthma, 97 patients came from low-income households, 49 patients came from middle-income households, and 30 patients came from high-income households.
# 
# In Patients and Their Heart Disease Status Distributed by Income Status, of the 858 patients that reported no heart disease, 337 patients came from low-income households, 319 patients came from middle-income households, and 202 patients came from high-income households. Of the 77 patients that reported having heart disease, 43 patients came from low-income households, 20 patients came from middle-income households, and 14 patients came from high-income households.
# 
# In Patients and Their Diabetes Status Distributed by Income Status, of the 858 patients that reported no diabetes, 337 patients came from low-income households, 319 patients came from middle-income households, and 202 patients came from high-income households. Of the 77 patients that reported having diabetes, 43 patients came from low-income households, 20 patients came from middle-income households, and 14 patients came from high-income households.

```

\newpage

### Results

In visualization 1, we can see that of those in the study population that did smoke, smoked an average of 13 cigarettes per day and have an average 21 pack years. Considering that 20-40 pack years are categorized as moderate smokers, these groups have higher odds in health outcomes and there could be effect modification on other confounders that we have not identified within the study.

Visualization 2 is an outcome table that portrays in our study population, those who were social smokers were about 75% of the study population, which may allude to a need for improved community public health intervention to reduce the prevalence on social smoking. Our outcomes table also includes prevalence percentages for the health outcomes of asthma, heart disease and other mental illnesses. If we wanted to investigate further, then we could you these percentages if we wanted to see if they are statistically significant enough for this study population.

Lastly, in visualization 3 when comparing all bar-graphs of health outcomes off asthma, heart disease, and diabetes stratified by income status, we can observe that the study population has a larger proportion of low-income patients, therefore we cannot assume that those with health outcomes are disproportionately low-income households. In patients who have heart disease, it is observed that they are the exact same patients that have diabetes, which we can interpret that heart disease and diabetes have some biological correlation. 

### Discussion

Based on our results, it seems that income is a considerable mediator for our health outcomes of interest related to smoking. The majority of people who reported adverse health outcomes were in the lowest income group, and the least amount of people reporting negative health outcomes were in the highest income group. For each negative health outcome, those in the lower income bracket were more likely to report the outcome than those in the middle income group. Those in the middle income group were more likely to report the negative health outcome than those in the highest income group. 

Additionally, we found that a person reporting average, below average, or above average pack years does not strongly correlate with disease outcome. Those who reported above-average pack years were no more likely to report asthma (19.9%)  or mental health issues (18.3%)  than those with below average pack years (19.9% and 18.8%). However, those with above average pack years were more likely to report diabetes (12.7%) than those with below average pack years (6%) and more likely to report heart disease (12.7%) than those with below average pack years (6%). 
	
Based on these findings, our recommendation for CDPH would be to prioritize quitting and smoking prevention campaigns and programs among lower income communities. We would not recommend campaigns to decrease smoking (and therefore pack years) since pack years did not seem to have as large of an effect on disease outcome. This recommendation for interventions targeted towards certain populations is  aimed at attempting  to reduce health disparities since based on our results, those in lower income groups are more likely to report negative health outcomes than those in higher income groups. 

