---
title: "Group AA Milestone 4"
output:
  pdf_document: default
  word_document: default
date: '2022-11-18'
group members: Victoria, Danni, Maddy
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r importing demographic data, include=FALSE}
#import CA Smoker data set
ca_smoker_info <- read_csv("~/PHW251_Fall2022/phw251_projectdata/ca_csc_smoker_data.csv")

ca_smoker_info$HOWMANY <- as.character(ca_smoker_info$HOWMANY)

#tidying data
ca_smoker_selected <- ca_smoker_info %>% select(c(psraid,smokstat,HOWMANY, SMOK6NUM, SMOK6UNI)) %>% 
rename(ID = psraid, smoking_status = smokstat, howmany = HOWMANY, smok6num = SMOK6NUM, smok6uni = SMOK6UNI) 

ca_smoker_selected$howmany <- as.numeric(ca_smoker_selected$howmany)
ca_smoker_selected$smok6num <- as.numeric(ca_smoker_selected$smok6num)

ca_smoker_selected <- ca_smoker_selected %>%
  mutate(pack_year = howmany/20*smok6num)
```

```{r importing outcome data, include=FALSE}
#import CA smoker disease outcome and race data set
ca_outcome_race <- read_csv("~/PHW251_Fall2022/phw251_projectdata/ca_csc_outcome_race_data.csv")

#tidying data
ca_outcome_race_selected <- ca_outcome_race %>% select(c(ID, SOCIAL, ASTHMA, 
                                                         HEARTDIS, DIABETES, 
                                                         OTHMENILL, INCOME, 
                                                         race01, race02, race03, 
                                                         race04, race05, race06, 
                                                         race07, race08, race09, 
                                                         race10, race11, race12, 
                                                         race13,race14, race15)) %>%
rename (social = SOCIAL, asthma = ASTHMA, heartdis = HEARTDIS, diabetes = DIABETES, 
        othmenill = OTHMENILL, income = INCOME)

```

```{r, include=FALSE}
#remove "DIS" & "STAT" in the "ID" column 
ca_outcome_race_selected$ID <- gsub("DIS","",as.character(ca_outcome_race_selected$ID))
ca_outcome_race_selected$ID <- gsub("STAT","",as.character(ca_outcome_race_selected$ID))

#joining two data sets by participant's unique ID
ca_smoker_outcome <- merge(x = ca_smoker_selected, y = ca_outcome_race_selected, by = "ID")

```


```{r create single race variable, include=FALSE}

#use mutate to combine 15 binary race columns into one categorical variable called "race"

ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(race = case_when(race01 == "Yes" ~ "White",
                          race02 == "Yes" ~ "Black",
                          race03 == "Yes" ~ "Japanese",
                          race04 == "Yes" ~ "Chinese",
                          race05 == "Yes" ~ "Filipino",
                          race06 == "Yes" ~ "Korean",
                          race12 == "Yes" ~ "Vietnamese",
                          race07 == "Yes" ~ "Other Asian Pacific Islander",
                          race08 == "Yes" ~ "American Indian Alaska Native",
                          race09 == "Yes" ~ "Mexican",
                          race10 == "Yes" ~ "Hispanic or Latino",
                          race11 == "Yes" ~ "Other",
                          race13 == "Yes" ~ "Asian Indian",
                          race14 == "Yes" ~ "Refused",
                          race15 == "Yes" ~ "Don't Know"))

#drop leftover binary race columns 

ca_smoker_outcome <- select(ca_smoker_outcome,-race01,
                                 -race02,
                                 -race03,
                                 -race04,
                                 -race05,
                                 -race06,
                                 -race07,
                                 -race08,
                                 -race09,
                                 -race10,
                                 -race11,
                                 -race12,
                                 -race13,
                                 -race14,
                                 -race15)
                            

```

```{r, include=FALSE}
#identify types of each data element
str(ca_smoker_outcome)

#convert data types to appropriate type in new column such as as.factor; as.numeric; as.character
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(new_howmany = as.numeric(howmany)) %>%
  mutate(new_smoking_status= as.factor(smoking_status)) %>% 
  mutate(new_smok6num = as.numeric(smok6num)) %>% 
  mutate(new_smok6uni = as.factor(smok6uni)) %>%
  mutate(new_pack_year = as.numeric(pack_year)) %>% 
  mutate(new_social = as.factor(social)) %>% 
  mutate(new_asthma = as.factor(asthma)) %>% 
  mutate(new_heartdis = as.factor(heartdis)) %>% 
  mutate(new_diabetes = as.factor(heartdis)) %>% 
  mutate(new_othmenill = as.factor(othmenill)) %>% 
  mutate(new_income = as.factor(income))

summary(ca_smoker_outcome$new_howmany)
```

```{r explore smoking status variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(smoking_status))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(smoking_status))

#tabulate smoking status
table(ca_smoker_outcome$smoking_status)
```


```{r explore race variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(race))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(race))

#tabulate smoking status
table(ca_smoker_outcome$race)
```

```{r explore income variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(income))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(income))

#tabulate income
table(ca_smoker_outcome$income)
```

```{r explore heart disease outcome variable, include=FALSE}
#number of unique categories 
ca_smoker_outcome %>% summarize(n_distinct(heartdis))

#names of unique categories 
ca_smoker_outcome %>% summarize(unique(heartdis))

#tabulate income
table(ca_smoker_outcome$heartdis)
```
```{r explore pack years variable, include=FALSE}

#look at minimum, median, mean, maximum, and # of NAs in pack year
summary(ca_smoker_outcome$new_pack_year)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)

is.na(ca_smoker_outcome) <- ca_smoker_outcome == "(DO NOT READ) Don't know"
is.na(ca_smoker_outcome) <- ca_smoker_outcome == "(DO NOT READ) Refused"
is.na(ca_smoker_outcome) <- ca_smoker_outcome == "NA/Not Applicable"

#subsetting rows  and columns to include only "new" columns, rows include NA 

str(ca_smoker_outcome)
ca_smoker_outcome <- select(ca_smoker_outcome, 
                            c('ID', 'new_smoking_status', 'new_howmany', 
                              'new_smok6num', 'new_smok6uni', 'new_pack_year',
                              'new_social', 'new_asthma', 'new_heartdis',
                              'new_diabetes', 'new_othmenill', 'new_income'))
```

```{r, include=FALSE}
library(stringr)
ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(smok_daily = str_c(ca_smoker_outcome$new_smok6num," ",
                            ca_smoker_outcome$new_smok6nuni)) 
```

```{r, include=FALSE}
library(dplyr)
summary(ca_smoker_outcome$new_pack_year)
ca_smoker_outcome <- ca_smoker_outcome %>% 
  mutate(pack_year_avg = mean(ca_smoker_outcome$new_pack_year, na.rm = TRUE))
ca_smoker_outcome$pack_year_avg <- round(ca_smoker_outcome$pack_year_avg ,digit= 0)
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(pack_year_avg_level = case_when
                                                  (new_pack_year > pack_year_avg ~ "above average", 
                                                    new_pack_year < pack_year_avg ~ "below average",
                                                    new_pack_year == pack_year_avg ~ "average"))

```

```{r, include=FALSE}
unique(ca_smoker_outcome$new_income)
ca_smoker_outcome <- ca_smoker_outcome %>% mutate(income_levels = case_when( 
  new_income %in% c("$20,000 or less", "$20,001 to $30,000") ~ "Low Income",
new_income %in% c("$30,001 to $50,000", "$50,001 to $75,000") ~ "Middle Income",
new_income %in% c("$75,001 to $100,000", "$100,001 to $150,000", "Over $150,000") ~ "High Income")) 
```

```{r, include=FALSE}
table(ca_smoker_outcome$new_smoking_status, useNA = "always")
table(ca_smoker_outcome$new_smok6uni, useNA = "always")
table(ca_smoker_outcome$new_social, useNA = "always")
table(ca_smoker_outcome$new_asthma, useNA = "always")
table(ca_smoker_outcome$new_heartdis, useNA = "always")
table(ca_smoker_outcome$new_diabetes, useNA = "always")
table(ca_smoker_outcome$new_othmenill, useNA = "always")
table(ca_smoker_outcome$new_income, useNA = "always")
```

```{r other new variable brainstorm, include=FALSE}

head(ca_smoker_outcome)

#if in years, leave as is
#if in months, divide by 12
#could add new column, conditional = 1 if units is in years, equals 12 if units is in months
#then create another new column for daily smoking length in years
#then update formula for pack years to reflect this

test_variable_df <- ca_smoker_outcome %>%
  mutate(temp_var = case_when(new_smok6uni=="Months" ~ 12,
                              new_smok6uni=="Years" ~ 1, 
                              new_smok6uni=="Days" ~ 365))

test_variable_df <- test_variable_df %>% 
  mutate(new_smok6num_inyears = new_smok6num/temp_var)

```

\newpage

# Milestone 4: Visualizations

## Visual 1: Table One of Pack Years and Health Outcomes
```{r}
library(kableExtra)
library(tableone) #load table one library

#build new dataset with only necessary variables
visual3_df <- test_variable_df %>% 
  select(new_asthma, new_diabetes, new_heartdis, new_pack_year, new_othmenill, pack_year_avg_level) %>%
  rename(Asthma=new_asthma,
         "Above/Below Average Pack Years" = pack_year_avg_level,
         Diabetes=new_diabetes, 
         "Heart Disease"=new_heartdis, 
         "Pack Years"=new_pack_year, 
         "Other Mental Health" =new_othmenill)
#create table one
visual3table <- CreateTableOne(data=visual3_df,
                               vars=c("Pack Years", "Asthma", "Diabetes", "Heart Disease", "Other Mental Health"),
                               factorVars = c("Asthma", "Diabetes", "Heart Disease", "Other Mental Health", "Above/Below Average Pack Years"),
                               strata="Above/Below Average Pack Years")
kable(print(visual3table, showAllLevels=TRUE), caption="Pack Years and Health Outcomes Stratified by Above/Below Average Pack Years")

```
### Interpretation: 
This table is a "table one": a common table of descriptive statistics for a study sample. This table one displays descriptive statistics for our health outcomes of interest, as well as the pack years variable. All are stratified by one of our factor variables, "above/below pack years average", which displays
whether the participant's pack years calculated in Milestone 3 were above, at, or below the average pack years of the entire sample. 

\newpage

## Visual 2: Outcome and Income Cross Tabulations 

```{r, include=FALSE}
library(janitor)
library(kableExtra)
library(tidyverse)
```

```{r}
# For asthma status data set
a <- test_variable_df %>%
  tabyl(new_asthma, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Asthma_status = new_asthma)

a <- a[, c("Asthma_status", "High Income", "Middle Income", "Low Income", "Total")]

a %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 1, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Asthma Status x Income Level", 5))
```

### Interpretation: 
Of the 176 patients with asthma, 30 were reported as the high-income group, 49 as the middle-income group, and 97 as the low-income group. 
Of the 762 study participants without asthma, 187 were reported as the high-income group, 290 as the middle-income group, and 285 as the low-income group.  

```{r}
# For heart disease status data set
b <- test_variable_df %>%
  tabyl(new_heartdis, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Heartdis_status = new_heartdis)

b <- b[, c("Heartdis_status", "High Income", "Middle Income", "Low Income", "Total")]

b %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 1, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Heart Disease Status x Income Level", 5))
```
Danni: For the "(DO NOT READ) Don't know" row, the level of variable factor is still there, not quite sure how to remove it

### Interpretation:
Of the 77 patients with heart disease, 14 were reported as the high-income group, 20 as the middle-income group, and 43 as the low-income group. 
Of the 858 study participants without heart disease, 202 were reported as the high-income group, 319 as the middle-income group, and 337 as the low-income group. 

```{r}
# For diabetes status data set
c <- test_variable_df %>%
  tabyl(new_diabetes, income_levels, show_na = FALSE) %>%
  adorn_totals(c("col", "row")) %>% 
  as.data.frame() %>% 
  rename(Diabetes_status = new_diabetes)

c <- c[, c("Diabetes_status", "High Income", "Middle Income", "Low Income", "Total")]

c %>% 
  kbl(align = "l") %>% 
  kable_styling(latex_options = "striped") %>% 
  add_header_above(c(" " = 1, "Income Level" = 3, " " = 1)) %>% 
  add_header_above(data.frame("Cross Tab of Diabetes Status x Income Level", 5))
```

### Interpretation:
Of the 77 patients with heart disease, 14 were reported as the high-income group, 20 as the middle-income group, and 43 as the low-income group. 
Of the 858 study participants without heart disease, 202 were reported as the high-income group, 319 as the middle-income group, and 337 as the low-income group. 

\newpage

## Visual 3: Of those who do have health outcomes, how are they distributed by income status. How are those that do not have health outcomes distributed by income?

```{r, include=FALSE}
library(ggplot2)
# Changed New variables to factors from most recent data set 
test_variable_df$pack_year_avg_level <- as.factor(test_variable_df$pack_year_avg_level)
test_variable_df$income_levels <- factor(test_variable_df$income_levels)
test_variable_df$smok_daily <- as.numeric(test_variable_df$smok_daily)
str(test_variable_df)

#Adding levels to Income Level 
test_variable_df$income_levels <- ordered(test_variable_df$income_levels, levels = c("Low Income", "Middle Income", "High Income"))
levels(test_variable_df$income_levels)

#Adding levels to pack year average level
test_variable_df$pack_year_avg_level <- ordered(test_variable_df$pack_year_avg_level, levels = c("below average", "average", "above average"))
levels(test_variable_df$pack_year_avg_level)
```
\newpage

## Asthma data set and ggplot

```{r}
asthma_by_income <- test_variable_df %>% select(new_asthma, income_levels) %>% 
  group_by(new_asthma, income_levels) %>%  mutate(asthma_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

A <- ggplot(data = na.omit(asthma_by_income), aes(x = new_asthma, y= asthma_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
    scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Asthama", y = "# of Patients",
       title = "Patients and Their Asthma Status Distributed by Income Status", 
       caption = "Note: Patients who did not respond were excluded")
```

## Heart Disease data set and ggplot

```{r}
heartdis_by_income <- test_variable_df %>% select(new_heartdis, income_levels) %>% 
  group_by(new_heartdis, income_levels) %>%  mutate(heartdis_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

B <- ggplot(data = heartdis_by_income, aes(x = new_heartdis, y= heartdis_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
  scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Heart Disease", y = "# of Patients", 
       title = "Patients and Their Heart Disease Status Distributed by Income Status",
       caption = "Note: Patients who did not respond were excluded")

```

## Diabetes data set and ggplot

```{r}
diabetes_by_income <- test_variable_df %>% select(new_diabetes, income_levels) %>% 
  group_by(new_diabetes, income_levels) %>%  mutate(diabetes_count = n()) %>% 
  distinct(income_levels,.keep_all = TRUE) %>% na.omit()

C<- ggplot(data = diabetes_by_income, aes(x = new_diabetes, y= diabetes_count)) +
  geom_bar(aes(fill = income_levels), stat = "identity", position = position_dodge()) +
  scale_fill_discrete(name = "Income Level") + 
  labs(x= "Prevalence of Diabetes", y = "# of Patients", 
       title = "Patients and Their Diabetes Status Distributed by Income Status",
       caption = "Note: Patients who did not respond were excluded")
 
```

\newpage

## All dodged barcharts combined together - Asthma, Heart Disease and Diabetes - Victoria

```{r, include=FALSE}
# The following code allows download of ggpubr to utilize the ggarrange function to 
# combine all seperate ggplots together. 

if(!require(devtools)) install.packages("devtools")
devtools::install_github("kassambara/ggpubr")
install.packages("ggpubr")
library(ggpubr)
```
```{r}
# Hidden code allows download of ggpubr to utilize the ggarrange function to 
# combine all seperate ggplots together. 

figure_one <- ggarrange(A, B, C, ncol = 1, nrow = 3)
figure_one

```

\newpage

## Bonus Table from Milestone 3: Smoker Outcomes Distributions

```{r first table for outcomes distribution}
#load kable library
library(knitr)
library(kableExtra)

#calculate data for proportion table
summary(ca_smoker_outcome$new_social)
summary(ca_smoker_outcome$new_asthma)
summary(ca_smoker_outcome$new_heartdis)
summary(ca_smoker_outcome$new_diabetes)
summary(ca_smoker_outcome$new_othmenill)

#build dataframe
outcome_variable <- c("Social Smoker", "Asthma", "Heart Disease", "Diabetes", "Other Mental Illness")
yes_count_variable <- c(748, 191, 81, 81, 171)
no_count_variable <- c(241, 809, 916, 916, 820)
NA_count_variable <- c(11, 0, 3, 3, 9)
yes_prop_variable <- c("74.8%", "19.1%", "8.1%", "8.0%", "17.1%")
no_prop_variable <- c("24.1%", "80.9%", "91.6%", "91.6%", "82.0%")
NA_prop_variable <- c("1.1%", "0", "0.3%", "0.3%", "0.9%")

proportion_df <- data.frame(outcome_variable, yes_count_variable, 
                               no_count_variable, NA_count_variable,
                               yes_prop_variable, no_prop_variable,
                               NA_prop_variable)

#create table
outcomes_table <- kable(proportion_df, booktabs=T, align="l",
                        col.names=c("Condition", "Yes Count", 
                                    "No Count", "Not Applicable",
                                    "Percent Yes", "Percent No", 
                                    "Percent N/A"))
outcomes_table
```
### Interpretation:
This table is a set of descriptive statistics related to our variables of interest, and the prevalence of each outcome in our sample. 

\newpage

## Bonus Table from Milestone 3: Spread of Continuous Variables

```{r build descriptives table}

#get data
summary(ca_smoker_outcome$new_howmany)
summary(ca_smoker_outcome$new_pack_year)

#build dataframe
measure <- c("Average Cigarettes Smoked Per Day", "Pack Years")
minimum <- c(1, 0.10)
median <- c(12, 17)
mean <- c(13.89, 21.68)
maximum <- c(60, 120)
distribution_df <- data.frame(measure, minimum, median, mean, maximum)

#create table
distribution_table <- kable(distribution_df, booktabs=T, align="lcccc",
                        col.names=c("Measure", "Minimum Value", 
                                    "Median Value", "Mean", 
                                    "Maximum Value"))
distribution_table
```

### Interpretation:
This table displays the spread of the variables "Average Cigarettes Smoked per Day" and "Pack Years" in our sample. 

\newpage
.
\newpage

### Not for grading, but also including our data dictionary from Milestone 3 with adjustment recommended by Will, which helped with page runoff. Thank you for this suggestion!
```{r, include=FALSE}
library("kableExtra")
```

```{r}
variable_name <- c("ID", "new_smoking_status", "new_howmany", "new_smok6num", 
                   "new_somk6uni", "new_pack_year", "new_social", "new_asthma", 
                   "new_heartdis", "new_diabetes", "new_othmenill", "new_income", "smok_daily",
                   "pack_year_avg", "pack_year_avg_level", "income_levels")

data_type <- c("numeric", "character", "numeric", "numeric", "character", "numeric",
               "character", "character", "character", "character", "character", 
               "character", "numeric", "numeric", "character", "character")

description <- c("Participant identification number", 
                 "Current Smoking Status: Current Daily Smoker (Smoked >99 and smokes every day), 
                 Current Nondaily Smoker(Smoked >99 and smokes some days)", 
              "During the past 30 days, on the days that you did smoke, about 
              'HOWMANY' cigarettes did you usually smoke per day?", 
              "How long have you been smoking on a daily basis?", "The unit for smol6num variable", 
              "A pack-year is used to describe how many cigarettes smoked in a person's lifetime, 
              with a pack equal to 20 cigarettes", 
              "yes/no the participant identifies as a social smoker", 
              "yes/no the participant has asthma",
              "yes/no the participant has heart disease", 
              "yes/no the participant has diabetes",
              "yes/no the participant reports 'other mental illness'",
              "participant household income",
              "cigarettes smoked daily",
              "average pack years for entire cohort, created for our average variable",
              "indicator for whether participant is below, at, or above the average pack years",
              "categorical income levels defined using California household income data")

data_dictionary <- data.frame(variable_name, data_type, description)
data_dictionary <- data_dictionary %>% 
  rename(
    "Variable Name" = variable_name,
    "Data Type" = data_type,
    "Description" = description)

kable(data_dictionary, "latex") %>%
  kable_styling(full_width=TRUE)
```